---
title: "bixi_project"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = false}
#library(tidyverse)
library(broom)
#library(car)
library(ggplot2)
library(cowplot)
library(dplyr)
library(gridExtra)
```

# Exploratory Data Analysis

```{r}
bixi <- read.csv("bixi6_part1.csv")
```

```{r,fig.width=10, fig.height=4, message=FALSE, warning=FALSE,}

# Ensure variables exist
stopifnot("dur" %in% names(bixi))
bixi$logdur <- log(bixi$dur)

# Reusable minimalist theme
theme_clean <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(size = 0.2),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey35"),
    axis.title = element_text(face = "bold"),
    plot.caption = element_text(color = "grey40", size = 9, margin = margin(t = 6))
  )

# Helper to add median/mean lines with labels
add_center_lines <- function(p, xvar, data) {
  med <- median(data[[xvar]], na.rm = TRUE)
  mn  <- mean(data[[xvar]], na.rm = TRUE)
  p +
    geom_vline(xintercept = med, linetype = 2, linewidth = 0.6) +
    geom_vline(xintercept = mn,  linetype = 1, linewidth = 0.6, alpha = 0.8) +
    annotate("text", x = med, y = Inf, vjust = 1.5, label = "Median", size = 3.2) +
    annotate("text", x = mn,  y = Inf, vjust = 3.0, label = "Mean",   size = 3.2)
}

p_dur <- ggplot(bixi, aes(x = dur)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.9) +
  geom_density(linewidth = 0.9) +
  geom_rug(alpha = 0.15, length = unit(2.5, "pt")) +
  labs(
    title = "Trip duration (minutes)",
    subtitle = "Histogram with density overlay;",
    x = "Duration (minutes)", y = "Density",
    caption = "Visual check for right skew and potential variance stabilization via log transform."
  ) +
  theme_clean
p_dur <- add_center_lines(p_dur, "dur", bixi)

p_logdur <- ggplot(bixi[is.finite(bixi$logdur), ], aes(x = logdur)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.9) +
  geom_density(linewidth = 0.9) +
  geom_rug(alpha = 0.15, length = unit(2.5, "pt")) +
  labs(
    title = "Log trip duration",
    subtitle = "Histogram with density overlay; ",
    x = "log(Duration)", y = "Density",
    caption = "Log transform typically reduces right tail and improves normality for linear modeling."
  ) +
  theme_clean
p_logdur <- add_center_lines(p_logdur, "logdur", bixi)

plot_grid(p_dur, p_logdur, nrow = 1, labels = c("A", "B"))
```

```{r eda_prec_en, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}
# Ensure variables and build rainy-day intensity
stopifnot("precip" %in% names(bixi))
bixi$Wet      <- as.integer(bixi$prec > 0)
bixi$rain_int <- ifelse(bixi$prec > 0, log1p(bixi$prec), NA_real_)
bixi_wet      <- subset(bixi, Wet == 1 & is.finite(rain_int))

# Reuse theme and helper from above (theme_clean, add_center_lines)

p_prec <- ggplot(bixi_wet, aes(x = precip)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.9) +
  geom_density(linewidth = 0.9) +
  geom_rug(alpha = 0.15, length = unit(2.5, "pt")) +
  labs(
    title = "Precipitation on rainy days",
    subtitle = "Histogram with density overlay; ",
    x = "Precipitation (mm)", y = "Density",
    caption = "Raw precipitation often shows a heavy right tail on rainy days."
  ) +
  theme_clean
p_prec <- add_center_lines(p_prec, "precip", bixi_wet)

p_logprec <- ggplot(bixi_wet, aes(x = rain_int)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, alpha = 0.9) +
  geom_density(linewidth = 0.9) +
  geom_rug(alpha = 0.15, length = unit(2.5, "pt")) +
  labs(
    title = "log1p(precipitation) on rainy days",
    subtitle = "Histogram with density overlay;",
    x = "log(1 + precipitation)", y = "Density",
    caption = "The log1p transform compresses the heavy tail,."
  ) +
  theme_clean
p_logprec <- add_center_lines(p_logprec, "rain_int", bixi_wet)

plot_grid(p_prec, p_logprec, nrow = 1, labels = c("C", "D"))

```

```{r eda_prec_en, message=FALSE, warning=FALSE}

#install.packages("corrplot")
library(corrplot)

# Keep only numeric columns with non-zero variance
bixi_num <- bixi[sapply(bixi, is.numeric)]
bixi_num <- bixi_num[, sapply(bixi_num, function(x) sd(x, na.rm = TRUE) > 0)]

# Correlation matrix
corr_matrix <- cor(bixi_num, use = "pairwise.complete.obs")

# Plot
corrplot(
  corr_matrix,
  method = "color",                         # colored heatmap
  type = "full",                            # show full matrix
  addCoef.col = "black",                    # add correlation values
  number.cex = 0.6,                         # size of numbers
  tl.col = "black",                         # text color
  tl.srt = 45,                              # rotate x labels
  col = colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(200)
)

```

# Business questions

## Q1 --- Do weekend BIXI trips tend to be longer?

**Design.** We model trip duration using linear regression. Because
`dur` (minutes) is positive and typically right-skewed, we analyze
$\log(\text{dur})$ to improve normality and stabilize residual variance.

**Variables.** - `Weekend`: indicator equal to 1 for Saturday/Sunday
(`jj ∈ {6,7}`), 0 otherwise. - Controls (for adjusted inference "even
after adjusting"): month (`mm` as factor), borough (`arrondissement` as
factor), temperature (centered) and light nonlinearity, rainfall split
into "wet day" and intensity on wet days.

**Models:**

1\. *Crude:*

$\log(\text{dur}) = \beta_0 + \beta_1\,Weekend + \varepsilon$.

2\. *Adjusted:*

-   $X_1$: weekend indicator (binary).
-   $X_2$: centered temperature.
-   $X_3$: squared centered temperature (i.e., $X_3 = X_2^2$).
-   $X_4$: wet-day indicator (binary).
-   $X_5$: rain intensity on wet days (e.g.,
    $X_5=\log(1+\text{prec})\cdot X_4$).
-   $X_6,\dots,X_{M}$: month dummies (one omitted as reference).
-   $X_{M+1},\dots,X_{P}$: borough dummies (one omitted as reference).

The model we estimate is

$$
E\!\left(Y \mid X_1,\dots,X_P\right)
=
\beta_0
+ \beta_1 X_1
+ \beta_2 X_2
+ \beta_3 X_3
+ \beta_4 X_4
+ \beta_5 X_5
+ \sum_{j=6}^{M} \delta_j\, D^{(month)}_j
+ \sum_{k=1}^{K} \alpha_k\, D^{(borough)}_k
$$

**Hypothesis.** $H_0:\beta_1=0$ vs $H_1:\beta_1\neq 0$. We test this
with a *t*-test from `lm()`

```{r, message=FALSE, warning=FALSE}
# Load data
bixi <- read.csv("bixi6_part1.csv")

# Remove trips shorter than 1.5 minutes
#bixi <- subset(bixi, dur >= 1.5)

# Construct variables (base R)
bixi$Weekend <- as.integer(bixi$jj %in% c(6,7))
bixi$mm <- factor(bixi$mm)
bixi$arrondissement <- factor(bixi$arrondissement)
bixi$Wet <- as.integer(bixi$prec > 0)
bixi$temp_c <- bixi$temp - 20
bixi$logdur <- log(bixi$dur)
bixi <- bixi[is.finite(bixi$logdur), ]

# Models
m_q1_crude <- lm(logdur ~ Weekend, data = bixi)
```

**Temperature (centered and quadratic).**\
We include both a linear and a quadratic term for temperature. The
hypothesis is that the effect of temperature on trip duration is not
purely linear: durations may increase with warmer weather up to a
comfortable point (\~20--22 °C) and then decrease when it becomes too
hot. To capture this curvature, we model temperature as a centered
linear term (Temp_c = Temp − 20) and a squared term (Temp_c²).

**Rainfall (Wet and log-intensity).**\
We decompose rainfall into two effects:\
1. A binary indicator Wet (1 = day with any rainfall, 0 = dry day). This
captures whether trips are systematically shorter on rainy days compared
to dry days.\
2. A continuous intensity effect, modeled as $\log(1 + \text{prec})$ on
wet days. The log transformation smooths the influence of rainfall:
small differences matter when rain is light (0--5 mm), but additional
millimeters at higher levels have diminishing impact.

```{r, message=FALSE, warning=FALSE}
m_q1_adj <- lm(
  logdur ~ Weekend +
    temp_c + I(temp_c^2) +
    Wet + log(1 + precip):Wet +
    mm + arrondissement,
  data = bixi
)

```

```{r}
summary(m_q1_crude)
```

```{r}
summary(m_q1_adj)
```

We first fit a crude model with weekend indicator $X_1$.\
Test of $H_0:\beta_1=0$ vs $H_A:\beta_1\neq 0$ gives\
$\hat{\beta}_1=0.0738$, $p=0.0023$. Since $p < 0.01$, the effect is
significant.\
Weekend trips are therefore about $\exp(0.0738)-1 \approx 7.7\%$ longer
than weekday trips in average.

We then fit an adjusted model including temperature, rainfall, month and
borough dummies.\
The weekend effect remains significant with $\hat{\beta}_1=0.0678$,
$p=0.0055$.\
Again, since $p < 0.01$, we reject $H_0$. Weekend trips are about 7.0%
longer in this model in average.

**Diagnostics: Cheking Model Assumptions**
#### 1. **Residuals vs Fitted Values (Homoscedasticity)**
The residuals are fairly centered around zero, and no strong fan-shape pattern appears. There is some spread, but it looks roughly even across fitted values.
Homoscedasticity assumption seems acceptable. No major concern here.

```{r, message=FALSE, warning=FALSE}
# Residuals vs Fitted plot to check for homoscedasticity
plot(m_q1_adj$fitted.values, m_q1_adj$residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red", lwd = 2)
```

#### 2. **Q-Q Plot (Normality of Residuals)**
The points follow the line fairly well in the middle but deviate at the tails (especially on the far left and right).
The residuals are approximately normal, but there are some deviations in the tails. This means your p-values and confidence intervals are still reliable in large samples, but some caution is needed with outliers/extreme cases. The log transformation already improved things, so this is acceptable.

```{r, message=FALSE, warning=FALSE}
# Q-Q plot to check normality of residuals
qqnorm(m_q1_adj$residuals)
qqline(m_q1_adj$residuals, col = "red", lwd = 2)
```

#### 3. **Histogram of Residuals**
The histogram is roughly bell-shaped, though slightly skewed with a few extreme values.
Normality is fairly well satisfied, consistent with the Q-Q plot. No major violation.

```{r, message=FALSE, warning=FALSE}
# Histogram of residuals to check normality
hist(m_q1_adj$residuals, breaks = 30, main = "Histogram of Residuals",
     xlab = "Residuals", col = "lightblue", border = "black")
```

#### 4. **Fitted vs Actual Values**
The points follow the line reasonably well, though there is scatter. Not perfect, but expected for social/behavioral data like bike trips.
The model has reasonable predictive performance. It captures the main trend but doesn’t perfectly predict individual durations (which is expected).

```{r, message=FALSE, warning=FALSE}
# Plot Fitted values vs Actual values
plot(m_q1_adj$fitted.values, bixi$logdur, 
     xlab = "Fitted Values", ylab = "Actual Values", 
     main = "Fitted vs Actual Plot")
abline(0, 1, col = "red", lwd = 2)
```

#### 5. **VIF (Multicollinearity)**
All VIFs are <5, so there is no multicollinearity concern.

```{r, message=FALSE, warning=FALSE}
# Calculate VIF to check for multicollinearity
library(car)
vif(m_q1_adj)
```

**Conclusion.** Both crude and adjusted models confirm that weekend
trips are significantly longer at the 1% level. The weekend effect is
robust to adjustment for climate, seasonality, and borough differences.

## Q2 How do weather conditions affect trip duration? Is the impact of weather conditions further accentuated on weekends?

#### Test 1 --- "Climate block" (all weather terms jointly)

-   **Null:** $H_0:$ all weather coefficients = 0 (temperature linear +
    quadratic, wet-day, and rain intensity).
-   **Alt:** at least one weather coefficient $\neq 0$.
-   **Procedure:** compare the **reduced** model (remove all weather
    terms) to the **full** model via `anova(reduced, full)`.
-   **Decision:** if $p<0.01$, reject $H_0$ → weather is globally
    significant.

#### Test 2 --- "Temperature block"

-   **Null:** $H_0:$ temperature linear and quadratic = 0.
-   **Alt:** at least one temperature coefficient $\neq 0$.
-   **Procedure:** compare the model **without temperature terms** to
    the **full** model.
-   **Decision:** if $p<0.01$, reject $H_0$ → temperature is significant
    (conditional on controls and rainfall).

#### Test 3 --- "Rainfall block"

-   **Null:** $H_0:$ wet-day and rain-intensity coefficients = 0.
-   **Alt:** at least one rainfall coefficient $\neq 0$.
-   **Procedure:** compare the model **without rainfall terms** to the
    **full** model.
-   **Decision:** if $p<0.01$, reject $H_0$ → rainfall is significant
    (conditional on controls and temperature).

```{r}
## Full model
m_full <- lm(
  logdur ~ Weekend +
    temp_c + I(temp_c^2) +
    Wet + log(1 + precip):Wet +
    mm + arrondissement,
  data = bixi
)
summary(m_full)
```

```{r}
## 1) No climate
m_no_climate <- lm(
  logdur ~ Weekend + mm + arrondissement,
  data = bixi
)

anova(m_no_climate, m_full)
```

```{r}
## 2) No temperature
m_no_temp <- lm(
  logdur ~ Weekend +
    Wet + log(1 + precip):Wet +
    mm + arrondissement,
  data = bixi
)
anova(m_no_temp, m_full)

```

```{r}
## 3) No rain
m_no_rain <- lm(
  logdur ~ Weekend +
    temp_c + I(temp_c^2) +
    mm + arrondissement,
  data = bixi
)

anova(m_no_rain, m_full)
```

**(i) Climate block (all weather terms jointly).**\
- Null: $H_0:\beta_2=\beta_3=\beta_4=\beta_5=0$.\
- Test: compare full model with climate vs. reduced model without
climate.\
- Result: $F=2.64$, $p=0.032$.\
- Decision: since $p>0.01$, we fail to reject $H_0$.\
- Conclusion: at the 1% level, there is no evidence that weather as a
whole affects trip duration.

------------------------------------------------------------------------

**(ii) Temperature block (linear and quadratic).**\
- Null: $H_0:\beta_2=\beta_3=0$.\
- Result: $F\approx 0.0001$, $p=0.9999$.\
- Decision: clearly $p>0.01$. Fail to reject $H_0$.\
- Conclusion: temperature shows no effect on trip duration.

------------------------------------------------------------------------

**(iii) Rainfall block (wet-day indicator and rain intensity).**\
- Null: $H_0:\beta_4=\beta_5=0$.\
- Result: $F=5.27$, $p=0.0052$.\
- Decision: since $p<0.01$, we reject $H_0$.\
- Conclusion: rainfall significantly affects trip duration at the 1%
level.

------------------------------------------------------------------------

**Interpretation.**

Although the full weather block is not significant at
1%(p=0.032),rainfall alone is. considering the full weather block also
contains non- significant variables such as temperature, which dilute
the joint test statistic. Testing rainfall separately isolates the
effect and shows it is significant.

**Continuing to the second part of the question- Is the impact of
weather conditions further accentuated on weekends:**

Let $Y=\log(\text{dur})$. Define: - $X_1$: Weekend (1=weekend,
0=weekday). -$X_2$：tempc，-$X_3$=$\text{temp}_c^2$,- C: Wet (1 if any
rain, 0 if dry). - $X_5 = \log(1+\text{prec})\cdot X_4$ (rain intensity
on wet days). - $D^{(month)}_j$ (month dummies; one omitted) and
$D^{(borough)}_k$ (borough dummies; one omitted).

We estimate $$
\begin{aligned}
E(Y\mid X)&= \beta_0+ \beta_1 X_1+ \beta_2 X_2 + \beta_3 X_3+ \beta_4 X_4 + \beta_5 X_5 \\&\quad + \sum_j \delta_j D^{(\text{month})}_j+ \sum_k \alpha_k D^{(\text{borough})}_k \\&\quad + \underbrace{\gamma_1 X_1X_2 + \gamma_2 X_1X_3}_{\text{Weekend}\times\text{Temperature}}+ \underbrace{\gamma_3 X_1X_4 + \gamma_4 X_1X_5}_{\text{Weekend}\times\text{Rain}} \, .
\end{aligned}
$$

**1.Temperature margin effect (weekday vs weekend)**

$$
\begin{aligned}\left.\frac{\partial E(Y\mid X)}{\partial X_2}\right|_{X_1=0}&= \beta_2 + 2\beta_3 X_2 \quad \text{(Weekday)},\\[2pt]\left.\frac{\partial E(Y\mid X)}{\partial X_2}\right|_{X_1=1}&= (\beta_2+\gamma_1) + 2(\beta_3+\gamma_2) X_2 \quad \text{(Weekend)}.\end{aligned}
$$

**2.Rainfall margin effect (weekday vs weekend).**

\- **Weekdays (** $X_1=0$ ): wet jump $=\beta_4$; intensity slope
$=\beta_5$.

\- **Weekends (** $X_1=1$ ): wet jump $=\beta_4+\gamma_3$; intensity
slope $=\beta_5+\gamma_4$.

On the original scale, percent changes are approximated by
$(e^{\hat\theta}-1)\times 100\%$.

------------------------------------------------------------------------

#### Hypotheses (at $\alpha=0.01$)

**Global "weekend modifies temp/rain" test** $$
H_0:\ \gamma_1=\gamma_2=0
\quad \text{vs} \quad
H_A:\ \text{at least one of }\{\gamma_1,\gamma_1\}\neq 0.
\\
H_0:\ \gamma_3=\gamma_4=0
\quad \text{vs} \quad
H_A:\ \text{at least one of }\{\gamma_3,\gamma_4\}\neq 0.
$$

(Use an F-test comparing the model **with** vs **without** the two
interaction terms.)

------------------------------------------------------------------------

```{r}
#full original model
m_no_int <- lm(logdur ~ Weekend + temp_c + I(temp_c^2) +
                   Wet + log(1+precip):Wet +
                   factor(mm) + arrondissement, data=bixi)
summary(m_temp_no_int)
```

```{r}
# with Weekend×Temp
m_temp_int <- lm(logdur ~ Weekend*(temp_c + I(temp_c^2)) +
                   Wet + log(1+precip):Wet +
                   factor(mm) + arrondissement, data=bixi)
```

```{r}

# block F-test: H0: gamma1=gamma2=0
anova(m_no_int, m_temp_int)  

```

**Weekend × Temp interactions.**\
The F-test comparing models with and without the Weekend×rain
interactions yields $F=0.21$ with $p=0.8044$. At the 1% significance
level, we fail to reject $H_0$.

```{r}
## With Weekend × rain interactions
m_rain_int <- lm(
  logdur ~ Weekend +
           temp_c + I(temp_c^2) +
           Weekend*(Wet + log(1+precip):Wet) +   # <-- Weekend×Rain interactions
           factor(mm) + arrondissement,
  data = bixi
)
summary(m_rain_int)
```

```{r}
## Global interaction block (γ3=γ4=0?) — 2 df}
anova(m_no_int, m_rain_int)   # decision at alpha = 0.01
```

**Weekend × rain interactions.**\
The F-test comparing models with and without the Weekend×rain
interactions yields $F=0.2785$ with $p=0.7569$. At the 1% significance
level, we fail to reject $H_0$.

**Interpretation.**\
There is no statistical evidence that the effect of rainfall (either the
wet-day dummy or rainfall intensity) nor Temperature differs between
weekends and weekdays. The negative effect of rainfall intensity on trip
duration appears stable across both weekdays and weekends.

```{r}
# Predictions across temperature by weekend status
newdat <- expand.grid(
  Weekend = c(0,1),
  temp_c  = seq(min(bixi$temp_c), max(bixi$temp_c), length.out = 100),
  Wet = 0, precip = 0,  # dry day for clarity
  mm = factor(5, levels = sort(unique(bixi$mm))),            # baseline month
  arrondissement = levels(bixi$arrondissement)[1]            # baseline borough
)
newdat$pred <- predict(m_no_int, newdata = newdat, se.fit = TRUE)
newdat$fit  <- newdat$pred$fit
newdat$se   <- newdat$pred$se.fit
newdat$lwr  <- newdat$fit - 1.96*newdat$se
newdat$upr  <- newdat$fit + 1.96*newdat$se

p_temp <-
ggplot(newdat, aes(temp_c, fit, color = factor(Weekend), fill = factor(Weekend))) +
  geom_line() +
  geom_ribbon(aes(ymin=lwr, ymax=upr), alpha=.15, color = NA) +
  labs(x = "Centered temperature (°C)", y = "Predicted log duration",
       color = "Weekend", fill = "Weekend",
       title = "Temperature effect on trip duration: Weekend vs. Weekday") +
  theme_minimal()

```

```{r}
wetdat <- expand.grid(
  Weekend = c(0,1),
  Wet = 1,
  precip = seq(0, quantile(bixi$precip[bixi$precip>0], .95, na.rm=TRUE), length.out=80),
  temp_c = 0, mm = factor(5, levels = sort(unique(bixi$mm))),
  arrondissement = levels(bixi$arrondissement)[1]
)
wetdat$log1p_prec <- log1p(wetdat$precip)
wetdat$pred <- predict(m_no_int, newdata = transform(wetdat, `log(1+precip)`=log1p_prec), se.fit=TRUE)
wetdat$fit  <- wetdat$pred$fit
wetdat$se   <- wetdat$pred$se.fit
wetdat$lwr  <- wetdat$fit - 1.96*wetdat$se
wetdat$upr  <- wetdat$fit + 1.96*wetdat$se

p_rain <-ggplot(wetdat, aes(precip, fit, color=factor(Weekend), fill=factor(Weekend))) +
  geom_line() +
  geom_ribbon(aes(ymin=lwr, ymax=upr), alpha=.15, color=NA) +
  labs(x="Rainfall (mm) on wet days", y="Predicted log duration",
       color="Weekend", fill="Weekend",
       title="Rainfall effect on trip duration (wet days): Weekend vs. Weekday") +
  theme_minimal()

```

```{r}
grid.arrange(
  p_temp + theme(legend.position = "bottom"),
  p_rain + theme(legend.position = "bottom"),
  ncol = 1,nrow=2
)
```

**Overall conclusion.**

Controlling for month and borough fixed effects, the rainfall block is
significant (α=0.01), while the temperature block is not.
Added Weekend×Rain and Weekend×Temperature interactions are not
significant, indicating no differential weather effect by weekend status
as shown on the figure above.

## Question 3 --- Are there any seasonal effects on trip duration?

We test whether **month indicators** are jointly significant in the
linear model for $Y=\log(\text{dur})$, controlling for weekend, weather,
and borough.

**Null and alternative (seasonality block).**\
Let $\{\delta_6,\delta_7,\delta_8,\delta_9,\delta_{10}\}$ be the
coefficients for months June--October (May is the reference). We test $$
H_0:\; \delta_6=\delta_7=\delta_8=\delta_9=\delta_{10}=0
\quad\text{vs}\quad
H_A:\; \text{at least one }\delta_j\neq 0,
$$ at significance level $\alpha=0.01$.

We compare a model **without** month dummies to the **full** model (with
month dummies) using a nested-model **F-test**.

```{r}
## Full model used in Q2 (includes month dummies)
m_full <- lm(
  logdur ~ Weekend +
    temp_c + I(temp_c^2) +
    Wet + log(1 + precip):Wet +
    mm + arrondissement,
  data = bixi
)

## Reduced model without month dummies
m_no_month <- lm(
  logdur ~ Weekend +
    temp_c + I(temp_c^2) +
    Wet + log(1 + precip):Wet +
    arrondissement,
  data = bixi
)

## Block F-test: are months jointly significant?
anova(m_no_month, m_full)
```

The block F-test for month dummies yields $F=4.79$, $p=0.00022$.\
At $\alpha=0.01$, we reject $H_0$ and conclude that there are
significant seasonal effects on trip duration.
